---
title: "In-Class Session 4: Applied Regression Workshop"
author: "Econometrics I"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(93)
library(fixest)
```

This is the hands-on session: everything here uses the Cornwell--Rupert wage data
from the lecture. (Yes, it's a labor example --- it's the canonical one for regression
mechanics; your assignments use data closer to your fields.)

# Reproduce: the Mincer regression

```{r}
wages <- read.csv("cornwell-rupert.csv")
names(wages) <- trimws(names(wages))
wages$EXP2 <- wages$EXP^2
m1 <- feols(LWAGE ~ ED + EXP + EXP2 + FEM + UNION, data = wages)
summary(m1)
```

Interpretation check: the coefficient on `ED` is in *log points* per year of
schooling. For small values these are approximately percent changes; exactly,
a coefficient $b$ means a $100(e^{b}-1)\%$ change.

```{r}
b <- coef(m1)["ED"]
c(log_points = unname(b), exact_pct = unname(100 * (exp(b) - 1)))
```

# Reproduce: interactions with `i()` and heterogeneous returns

Do returns to education differ by sex?

```{r}
m2 <- feols(LWAGE ~ ED * FEM + EXP + EXP2 + UNION, data = wages)
summary(m2)
```

# Reproduce: the delta method by hand

Suppose we want the standard error of the *experience level that maximizes wages*,
$\theta = -\beta_{EXP} / (2\beta_{EXP^2})$ --- a nonlinear function of two
coefficients. The delta method: $\widehat{Var}(\hat\theta) = \nabla g' \, \hat V \, \nabla g$.

```{r}
m3 <- feols(LWAGE ~ ED + EXP + EXP2, data = wages)
b  <- coef(m3); V <- vcov(m3)
b1 <- b["EXP"]; b2 <- b["EXP2"]
theta <- -b1 / (2 * b2)
grad  <- c(EXP = -1 / (2 * b2), EXP2 = b1 / (2 * b2^2))
se_theta <- sqrt(t(grad) %*% V[c("EXP", "EXP2"), c("EXP", "EXP2")] %*% grad)
c(theta = unname(theta), se = as.numeric(se_theta))
```

# Reproduce: the bootstrap alternative

```{r}
B <- 499
boot_theta <- replicate(B, {
  idx <- sample(nrow(wages), replace = TRUE)
  bb  <- coef(feols(LWAGE ~ ED + EXP + EXP2, data = wages[idx, ]))
  -bb["EXP"] / (2 * bb["EXP2"])
})
c(boot_se = sd(boot_theta))
quantile(boot_theta, c(0.025, 0.975))
```

Compare to the delta-method interval --- close, with no calculus required.

# In-class problem 1: a union premium for whom?

**(a)** Add a `UNION` $\times$ `ED` interaction to the Mincer regression. What is the
estimated return to education for union vs. non-union workers?
**(b)** Test the hypothesis that the returns are equal (i.e., the interaction is zero)
--- and then test the *joint* hypothesis that union status doesn't matter at all
(level and interaction both zero).
**(c)** One sentence: why are these two different questions?

```{r problem1}
# your code here
```

# In-class problem 2: attenuation, live

**(a)** Create a noisy version of education: `ED_noisy = ED + rnorm(n, sd = 2)`.
Re-run the basic Mincer regression with `ED_noisy` in place of `ED`. What happened
to the coefficient?
**(b)** The reliability ratio is $\lambda = Var(ED) / (Var(ED) + 4)$. Compute it and
check that the noisy coefficient is roughly $\lambda$ times the clean one.
**(c)** Now try `sd = 6`. One sentence: in which direction does measurement error in a
regressor push your estimate, and can more data fix it?

```{r problem2}
# your code here
```
