---
title: "In-Class Session 3: Inference and Standard Errors"
author: "Econometrics I"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(93)
library(fixest)
library(MASS)
```

This notebook reproduces the standard-error comparisons from Lecture 3 and leaves
you two short problems.

# Reproduce: robust vs. classical under heteroskedasticity

A DGP where the error variance grows with $x$:

```{r}
n <- 2000
x <- rnorm(n)
y <- 1 + 2 * x + x^2 * rnorm(n)   # heteroskedastic errors
df <- data.frame(y, x)

m <- feols(y ~ x, data = df)
summary(m)                    # classical
summary(m, vcov = "hetero")   # HC1 "robust"
```

Same point estimate, different standard errors --- and here the classical one is too
small.

# Reproduce: clustering

Group-level shocks, 25 groups of 20:

```{r}
G <- 25; ng <- 20
g  <- rep(1:G, each = ng)
xg <- rnorm(G)[g] + rnorm(G * ng)       # x correlated within group
eg <- rnorm(G)[g] + rnorm(G * ng)       # error correlated within group
yg <- 1 + 2 * xg + eg
dfc <- data.frame(yg, xg, g)

mc <- feols(yg ~ xg, data = dfc)
rbind(classical = se(mc),
      robust    = se(summary(mc, vcov = "hetero")),
      cluster   = se(summary(mc, vcov = ~g)))
```

# Reproduce: the bootstrap in ten lines

```{r}
B <- 999
boot_slopes <- replicate(B, {
  idx <- sample(n, replace = TRUE)
  coef(lm(y ~ x, data = df[idx, ]))["x"]
})
c(boot_se = sd(boot_slopes), robust_se = se(summary(m, vcov = "hetero"))["x"])
quantile(boot_slopes, c(0.025, 0.975))   # percentile CI
```

# In-class problem 1: when two t-tests mislead

Simulate $y_i = 1 + 0.15 x_{1i} + 0.15 x_{2i} + \varepsilon_i$ with
$corr(x_1, x_2) = 0.95$, $\varepsilon \sim \mathcal{N}(0,1)$, $n = 300$.

**(a)** Run the regression. Look at the individual t-statistics on $x_1$ and $x_2$.
**(b)** Now test the *joint* hypothesis $\beta_1 = \beta_2 = 0$ (in `fixest`:
`wald(model, "x")` tests all coefficients matching `"x"`; or use an F-test comparing
to the intercept-only model).
**(c)** Two sentences: why can both individual tests fail to reject while the joint
test rejects decisively? Which picture from lecture is this?

```{r problem1}
# your code here
```

# In-class problem 2: how many clusters is enough?

Take the clustering DGP from above, but wrap it in a function of $G$ (keep total
$n = 500$ fixed: $G$ groups of $500/G$).

**(a)** For $G \in \{5, 10, 25, 50\}$, simulate 200 datasets each, and compute the
*coverage* of the 95% cluster-robust confidence interval for $\beta = 2$.
**(b)** What happens with very few clusters, and what does this imply about the
"cluster at the state level" advice when there are only a handful of states in your
data?

```{r problem2}
# your code here
```
