\documentclass[xcolor=pdftex,dvipsnames,table,mathserif,aspectratio=169]{beamer}
\usetheme{default}
\usetheme{metropolis}
\setbeamersize{text margin left=.3in,text margin right=.3in} 

%\usetheme{Darmstadt}
%\usepackage{times}
%\usefonttheme{structurebold}

\usepackage[english]{babel}
%\usepackage[table]{xcolor}
\usepackage{pgf,pgfarrows,pgfnodes,pgfautomata,pgfheaps}
\usepackage{amsmath,amssymb,setspace,centernot}
\usepackage[latin1]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{relsize}
\usepackage{pdfpages}
\usepackage[absolute,overlay]{textpos} 


\newenvironment{reference}[2]{% 
  \begin{textblock*}{\textwidth}(#1,#2) 
      \footnotesize\it\bgroup\color{red!50!black}}{\egroup\end{textblock*}} 

\DeclareMathSizes{10}{10}{6}{6} 

\begin{document}
\title{Program Evaluation(b):
Parametric Selection}
\author{Chris Conlon}
\institute{Applied Econometrics}
\date{\today}

\frame{\titlepage}

\begin{frame}
\frametitle{Example: Borjas (1987)}
\begin{itemize}
\item Consider two countries $(0/1)$ (source and host).
\begin{align*}
\ln w_0 &= \alpha_0 + u_0 \quad \mbox{ with } u_0 \sim N(0,\sigma_0^2) \mbox{ source country}\\
\ln w_1 &= \alpha_1 + u_1 \quad \mbox{ with } u_1 \sim N(0,\sigma_1^2) \mbox{ host country}
\end{align*}
\item Now we allow for migration cost of $C$ which he writes in hours: $\pi = \frac{C}{w_0}$.
\item Assume workers know everything; you only see $u_0$ \alert{OR} $u_1$ depending on country.
\item Correlation in earnings is $\rho=\frac{\sigma_{01}}{\sigma_{0} \sigma_{1}}$.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Example: Borjas (1987)}
\begin{itemize}
\item Workers will migrate if:
\begin{align*}
\left( \alpha_{1}-\alpha_{0}-\pi\right)+\left(u_{1}-u_{0}\right)>0
\end{align*}
\item Who migrates? Probability of migration. Define $\nu = u_1-u_0$.
\begin{align*}
P &=\operatorname{Pr}\left[\nu>\left(\alpha_{0}-\alpha_{1}+\pi\right)\right] =\operatorname{Pr}\left[\frac{\nu}{\sigma_{\nu}}>\frac{\left(\alpha_{0}-\alpha_{1}+\pi\right)}{\sigma_{\nu}}\right] \\
&=1-\Phi\left(\frac{\left(\alpha_{0}-\alpha_{1}+\pi\right)}{\sigma_{\nu}}\right) \equiv 1-\Phi(z)
\end{align*}
\item Higher $z$ $\rightarrow$ less migration.
\end{itemize}
\end{frame}


\begin{frame}
\frametitle{Example: Borjas (1987): How does selection work?}
Construct \alert{counterfactual wages} for workers in \alert{source} country for those who immigrate:
\begin{itemize}
\item For now ignore mean differences $\alpha_0 = \alpha_1 = \alpha$.
\begin{align*}
\mathbb{E} \left(w_{0} | \text { Immigrate }\right) &=\alpha+\mathbb{E}\left(u_{0} | \frac{\nu}{\sigma_{\nu}}>z\right) \\
&=\alpha+\sigma_{0} \cdot \mathbb{E}\left(\frac{u_{0}}{\sigma_{0}} | \frac{\nu}{\sigma_{\nu}}>z\right)
\end{align*}
\item Wages depend on:
\begin{enumerate}
\item Mean earnings in the source country
\item Both error terms $\left(u_{0}, u_{1}\right)$ through $\nu$
\item Implicitly, it also depends on the correlation between the error terms.
\end{enumerate}
\end{itemize}
\end{frame}


\begin{frame}
\frametitle{Example: Borjas (1987): How does selection work?}
\begin{itemize}
\item If everything is normal, we just run univariate regression $\mathbb{E}\left(u_{0} | \nu\right)=\frac{\sigma_{0 \nu}}{\sigma_{\nu}^{2}} \nu$:
\begin{align*}
\mathbb{E}\left(\frac{u_{0}}{\sigma_{0}} | \frac{\nu}{\sigma_{\nu}}\right) &= \frac{1}{\sigma_{0}}\cdot \frac{\sigma_{0 \nu}}{\sigma_{\nu}^{2}} \cdot \frac{\sigma_{\nu}^{2}}{\sigma_{\nu}^{2}} \cdot \nu  
=\frac{\sigma_{0 \nu}}{\sigma_{0} \sigma_{\nu}} \frac{\nu}{\sigma_{\nu}}  =\rho_{0 \nu} \frac{\nu}{\sigma_{\nu}} 
\end{align*}
\begin{align*}
\mathbb{E}\left(w_{0} | \text { Immigrate }\right) 
&=\alpha_{0}+\sigma_{0} \cdot \mathbb{E}\left(\frac{u_{0}}{\sigma_{0}} | \frac{\nu}{\sigma_{\nu}}>z\right) \\
 &=\alpha_{0}+\rho_{0 \nu} \cdot \sigma_{0} \cdot \mathbb{E}\left(\frac{\nu}{\sigma_{\nu}} | \frac{\nu}{\sigma_{\nu}}>z\right) \\
  &=\alpha_{0}+\rho_{0 \nu} \cdot \sigma_{0}\left(\frac{\phi(z)}{1-\Phi(z)}\right) 
\end{align*}
\item This hazard rate of the standard normal has a special name \alert{Inverse Mills Ratio} $\mathbb{E}[x | x > z]$.
\end{itemize}
\end{frame}


\begin{frame}
\frametitle{Example: Borjas (1987): How does selection work?}
\begin{itemize}
\item A similar expression for those who do immigrate:
\begin{align*}
\mathbb{E}\left(w_{1} | \text { Immigrate }\right) &=\alpha_{1}+\mathbb{E}\left(u_{1} | \frac{\nu}{\sigma_{\nu}}>z\right) \\
&=\alpha_{1}+\rho_{1 \nu} \sigma_{1}\left(\frac{\phi(z)}{\Phi(-z)}\right)
\end{align*}
\item We can re-write both expressions in terms of the \alert{Inverse Mills Ratio}
\end{itemize}
\end{frame}

\begin{frame}{Inverse Mills Ratio}
\begin{align*}
\mathbb{E}\left(w_{0} | \text { Immigrate }\right) &=\alpha_{0}+\rho_{0 \nu} \sigma_{0}\left(\frac{\phi(z)}{1-\Phi(z)}\right) \\
&=\alpha_{0}+\frac{\sigma_{0} \sigma_{1}}{\sigma_{\nu}}\left(\rho-\frac{\sigma_{0}}{\sigma_{1}}\right)\left(\frac{\phi(z)}{1-\Phi(z)}\right) \\
\mathbb{E}\left(w_{1} | \text { Immigrate }\right) &=\alpha_{1}+\rho_{1 \nu} \sigma_{1}\left(\frac{\phi(z)}{1-\Phi(z)}\right) \\
&=\alpha_{1}+\frac{\sigma_{0} \sigma_{1}}{\sigma_{\nu}}\left(\frac{\sigma_{1}}{\sigma_{0}}-\rho\right)\left(\frac{\phi(z)}{1-\Phi(z)}\right)
\end{align*}
Where $\rho=\sigma_{01} / \sigma_{0} \sigma_{1}$.
\end{frame}



\begin{frame}{Positive Hierarchical Sorting}
Let $Q_{0}=E\left(u_{0} | I=1\right), Q_{1}=E\left(u_{1} | I=1\right)$ (expected \alert{skill} of immigrants).
\begin{itemize}
\item Immigrants are positively selected and above average $(Q_0,Q_1) > 0$ and $\frac{\sigma_{1}}{\sigma_{0}}>1 \text { and } \rho>\frac{\sigma_{0}}{\sigma_{1}}$
\begin{itemize}
\item $\frac{\sigma_{1}}{\sigma_{0}}>1$ returns to ``skill'' are higher in host country.
\item $\rho>\frac{\sigma_{0}}{\sigma_{1}}$ correlation between valued skills in both counties is high (similar skills valued in both countries).
\end{itemize}
\item Best and brightest leave because returns to skill are too low in home country.
\end{itemize}
\end{frame}

\begin{frame}{Negative Hierarchical Sorting}
We swap the standard deviations:
\begin{itemize}
\item Immigrants are negatively selected and below average $(Q_0,Q_1) < 0$ and $\frac{\sigma_{1}}{\sigma_{0}}>1 \text { and } \rho>\frac{\sigma_{0}}{\sigma_{1}}$
\begin{itemize}
\item $\frac{\sigma_{0}}{\sigma_{1}}>1$ returns to ``skill'' are lower in host country.
\item $\rho>\frac{\sigma_{1}}{\sigma_{0}}$ correlation between valued skills in both counties is high (similar skills valued in both countries).
\end{itemize}
\item Compressed wage structure attracts the low skill types because it provides ``insurance'' or ``subsidizes'' low wage workers.
\end{itemize}
\end{frame}

\begin{frame}{Refugee/Superman Sorting?}
\begin{itemize}
\item Immigrants are below average at home and above average in host $(Q_0<0 ,Q_1>1)$ and $\frac{\sigma_{1}}{\sigma_{0}}>1$:
\begin{itemize}
\item $\rho<\min \left(\frac{\sigma_{1}}{\sigma_{0}}, \frac{\sigma_{0}}{\sigma_{1}}\right)$ being below average in source country makes you above average in host country.
\end{itemize}
\item You are a nerdy intellectual in a country that values physical labor, or are otherwise discriminated against in the labor market.
\end{itemize}
The missing (fourth) case:
\begin{itemize}
\item Mathematically impossible $\rho>\max \left(\frac{\sigma_{1}}{\sigma_{0}}, \frac{\sigma_{0}}{\sigma_{1}}\right)$
\end{itemize}
\end{frame}


\begin{frame}{Takeaway}
What can we learn here?
\begin{itemize}
\item Heckman won a Nobel Prize for his work on selection...
\item You need to know what an \alert{inverse Mills ratio is}
\item But today it is hard to get away with strong parametric assumptions (bivariate normal) on error terms.
\item Doing MLE with a fully normal model is not a terrible place to start sometimes
\begin{itemize}
\item Sometimes helpful to know how bad the selection problem might be.
\end{itemize}
\item R package is \texttt{sampleSelection} and see \url{https://rpubs.com/hacamvan/316839} and \url{https://cran.r-project.org/web/packages/sampleSelection/vignettes/selection.pdf}.
\end{itemize}
\end{frame}

\end{document}
