---
title: "In-Class Session 1: Probability, Operators, and the Big Theorems"
author: "Econometrics I"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(93)
```

This notebook reproduces the code from Lecture 1 and leaves you two short problems.
Run every chunk yourself. Nothing here requires more than base R and `MASS`.

# Reproduce: distributions from the slides

The normal CDF and PDF (the slides used `ggplot2`; base R's `curve` does fine):

```{r}
par(mfrow = c(1, 2))
curve(pnorm(x), from = -4, to = 4, main = "Standard Normal CDF", ylab = "F(x)")
curve(dnorm(x), from = -4, to = 4, main = "Standard Normal PDF", ylab = "f(x)")
```

# Reproduce: the Bayes / rare disease calculation

A test has a 5% false positive rate and a 0% false negative rate. If 1% of the
population has the disease, what is the probability a positive test means disease?

```{r}
posterior <- function(prevalence, fpr = 0.05, fnr = 0.00) {
  p_pos <- (1 - fnr) * prevalence + fpr * (1 - prevalence)
  (1 - fnr) * prevalence / p_pos
}
posterior(0.01)     # 1% prevalence
posterior(0.0005)   # 0.05% prevalence: "think horses, not zebras"
```

# Reproduce: the variance of a sample mean, from the operator rules

In lecture we derived $Var(\bar{X}_n) = \sigma^2/n$ using only linearity and the
variance rules. Let's watch the derivation be true:

```{r}
n <- 25; sigma2 <- 4
means <- replicate(2000, mean(rnorm(n, mean = 0, sd = sqrt(sigma2))))
c(simulated = var(means), predicted = sigma2 / n)
```

# Reproduce: LLN and CLT histograms

The Bernoulli sample-mean histograms from Lecture 1b:

```{r}
lln_plot <- function(n, p = 0.4) {
  means <- rbinom(1000, n, p) / n
  hist(means, xlim = c(0, 1), col = "beige", main = paste0("n = ", n),
       xlab = "Sample mean")
  abline(v = p, lwd = 3, col = "red")
}
par(mfrow = c(1, 3))
for (n in c(10, 100, 1000)) lln_plot(n)
```

```{r}
clt_plot <- function(n, p = 0.4) {
  z <- (rbinom(1000, n, p) / n - p) * sqrt(n)
  hist(z, xlim = c(-2, 2), col = "beige", freq = FALSE,
       main = paste0("standardized, n = ", n), xlab = "")
  curve(dnorm(x, sd = sqrt(p * (1 - p))), add = TRUE, lwd = 2, col = "red")
}
par(mfrow = c(1, 3))
for (n in c(10, 100, 1000)) clt_plot(n)
```

# In-class problem 1: diversification is a variance rule

Two assets have returns with standard deviation 10% each and correlation $\rho$.
From the operator rules, the equal-weighted portfolio has variance
$\sigma^2(1+\rho)/2$.

**(a)** Simulate 10,000 pairs of returns with $\rho = 0.9$ (use `MASS::mvrnorm`),
form the portfolio return, and compare its variance to the formula.
**(b)** Repeat for $\rho = -0.9$. What happened, and why is this "the entire logic
of diversification" in one number?

```{r problem1}
library(MASS)
# your code here
```

# In-class problem 2: total variance in a customer base

A subscription business has two customer segments. Segment A (30% of customers)
spends $\sim \mathcal{N}(80, 10^2)$ per month; segment B (70%) spends
$\sim \mathcal{N}(40, 20^2)$.

**(a)** Simulate 10,000 customers. Compute the total variance of spending.
**(b)** Compute the two pieces of the law of total variance --- the average
*within*-segment variance and the variance of the segment *means* --- and verify
they sum to (a).
**(c)** One sentence: if you could only report one of the two pieces to describe
"how different are our customers," which would you pick and why?

```{r problem2}
# your code here
```
