\documentclass[11pt]{article}
\usepackage{fullpage}
\usepackage[left=1in,top=1in,right=1in,bottom=1in,headheight=3ex,headsep=3ex]{geometry}

\newcommand{\blankline}{\quad\pagebreak[2]}

\title{Econometrics of Panel Data}\author{Chris Conlon}
\date{Spring 2019}

\usepackage[sc]{mathpazo}
\linespread{1.05}         % Palatino needs more leading (space between lines)
\usepackage[T1]{fontenc}

\usepackage[mmddyyyy]{datetime}% http://ctan.org/pkg/datetime
\usepackage{advdate}% http://ctan.org/pkg/advdate
\newdateformat{syldate}{\twodigit{\THEMONTH}/\twodigit{\THEDAY}}
\newsavebox{\THURSDAY}\savebox{\THURSDAY}{Tue}% Mon

\newcommand{\week}[1]{%
%  \cleardate{mydate}% Clear date
% \newdate{mydate}{\the\day}{\the\month}{\the\year}% Store date
  \paragraph*{\kern-2ex\quad #1, \syldate{\today}:}% Set heading  \quad #1
%  \setbox1=\hbox{\shortdayofweekname{\getdateday{mydate}}{\getdatemonth{mydate}}{\getdateyear{mydate}}}%
  \ifdim\wd1=\wd\THURSDAY
    \AdvanceDate[7]
  \else
    \AdvanceDate[7]
  \fi%
}



\usepackage{setspace}
\usepackage{multicol}
%\usepackage{indentfirst}
\usepackage{fancyhdr,lastpage}
\usepackage{url}
\pagestyle{fancy}
\usepackage{hyperref}
\usepackage{lastpage}
\usepackage{amsmath}
\usepackage{layout}   
\lhead{}
\chead{}
\rhead{\footnotesize Applied Econometrics- Fall 2019}
\lfoot{}
\cfoot{\small \thepage/\pageref*{LastPage}}
\rfoot{}

\usepackage{array, xcolor}
\usepackage{color,hyperref}
\definecolor{clemsonorange}{HTML}{EA6A20}
\hypersetup{colorlinks,breaklinks,
            linkcolor=clemsonorange,urlcolor=clemsonorange,
            anchorcolor=clemsonorange,citecolor=black}




\begin{document}


\maketitle

\blankline

\begin{tabular*}{.93\textwidth}{@{\extracolsep{\fill}}lr}


  E-mail: \texttt{cconlon@stern.nyu.edu} & Web: \href{http://chrisconlon.org}{\tt\bf chrisconlon.org/appliedmetrics}  \\

 Office Hours: Tues/Thurs 11-12 (or by appointment) &  Class Hours: Friday 1:00-4:00 \\


 Office: KMC 7-76  & Class Room: TBA \\
&  \\
\hline
\end{tabular*}

\vspace{10 mm}

\section*{Course Description}
This is a second Ph.D. course in applied econometrics though advanced undergraduates are welcome. The focus is on microeconometrics and panel data. It is a continuation of Prof. Scott's course, though once the basics are covered we will have more opportunity to explore topics related to student intrest.

\begin{description}
\item[Problem Sets:] I have designed the problem sets in R, though you are free to use whichever statistical software you would like. 
\end{description}

\section*{Books}
I will follow two main textbooks.
\begin{itemize}
\item Greene (2017). \textit{Econometric Analysis}. ISBN: 0134461363
\item Tibshirani, Hastie, Friedman (2016), \textit{The Elements of Statistical Learning}. ISBN: 0387848576. Available online at \url{https://web.stanford.edu/~hastie/Papers/ESLII.pdf}.
\end{itemize} 

\section*{Course Policy}
You are expected to attend every lecture and it is expected that you have done the reading BEFORE the class. This is a Ph.D. course which means you will be expected to read a lot on your own.
\subsection*{Grading Policy}
\begin{itemize}
\item 60\% of your grade will be performance on 6 problem sets (10\% each).
\item 30\% of your grade will be performance on the final exam.
\item 10\% of your grade will be participation in class.
\end{itemize}

\subsection*{Academic Dishonesty Policy}
Don't cheat. It is helpful to work with a partner on debugging code, but my expectation is that assignments are 100\% your own work (including computer code).

\newpage

\SetDate[01/02/2019]
\week{Week 01} Introduction to Time Series Data
\week{Week 02} Panel Data I : Fixed Effects, Random Effects, Clustering\\
\textit{PS 1 Due}\\
\week{Week 03} Maximum Likelihood and Duration Models\\
\week{Week 04} Discrete Choice and Multinomial Choice\\
\textit{PS 2 Due}
\week{Week 05} Bayesian Methods and MCMC\\
\week{Week 06} Panel Data II : Dynamic Panel,  Causal FE, Empirical Bayes, Hierarchical Models
\textit{PS 3 Due}
\week{Break} \textbf{SPRING BREAK (3/18-3/24)}
\week{Week 07} Treatment Effects and Potential Outcomes\\
\textit{PS 4 Due}
\week{Week 08} Treatment Effects and Potential Outcomes
\week{Week 09} Nonparametric Methods (Kernels, Nearest Neighbors, Bootstrap)\\
\textit{PS 5 Due}\\
\week{Week 10} Nonparametric Methods (Kernels, Nearest Neighbors, Bootstrap)
\week{Week 11} Machine Learning: Model Selection and Regularization (LASSO, RIDGE, PCA)\\
\week{Week 12} Topics based on interest: Duration Models, Dynamic Discrete Choice, Tree Models, Model Averaging Boosting/Bagging, etc.\\
\textit{PS 6 Due}
\week{Week 13} Topics based on interest (continued)





\end{document}