---
title: "In-Class Session 5: Causality, DAGs, and Experiment Design"
author: "Econometrics I"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(93)
```

This notebook reproduces the power calculations and the collider simulation from
Lecture 5, and leaves you two short problems.

# Reproduce: power and sample size for an A/B test

From the lecture: baseline conversion 2%, and we want to detect a lift to 2.5%
(a 25% relative lift) at $\alpha = 0.05$ with 80% power.

```{r}
power.prop.test(p1 = 0.02, p2 = 0.025, sig.level = 0.05, power = 0.80)
```

The required $n$ *per arm*. Now the sample-size/lift curve:

```{r}
lifts <- seq(0.0005, 0.005, by = 0.0005)
n_req <- sapply(lifts, function(d)
  power.prop.test(p1 = 0.02, p2 = 0.02 + d, sig.level = 0.05, power = 0.80)$n)
plot(100 * lifts / 0.02, n_req / 1000, type = "b", log = "y",
     xlab = "Detectable relative lift (%)", ylab = "n per arm (thousands, log scale)")
```

Small lifts are *brutally* expensive to detect --- this is the Lewis--Rao point about
advertising experiments.

# Reproduce: the collider from the DAG lecture

Talent and luck are independent. Funding requires enough of either:

```{r}
n      <- 100000
talent <- rnorm(n)
luck   <- rnorm(n)
funded <- (talent + luck > 1.5)   # top ~14% get funded

cor(talent, luck)            # population: ~0
cor(talent[funded], luck[funded])   # among the funded: strongly negative!
plot(talent[funded][1:2000], luck[funded][1:2000], pch = ".",
     xlab = "talent", ylab = "luck", main = "Funded startups only")
```

Nobody made talent and luck trade off --- *conditioning on the collider did*.

# In-class problem 1: the bad control, live

Simulate an advertising funnel: exposure to an ad is randomized,
`exposed ~ Bernoulli(0.5)`. Visiting the site depends on exposure and interest:
`visit = 1{0.5*exposed + interest + noise > 1}`, where `interest ~ N(0,1)`.
Purchases depend on exposure (through brand awareness) and interest:
`purchase = 0.1*exposed + 0.5*interest + error`.

**(a)** Regress `purchase` on `exposed` in the full sample. Since exposure is
randomized, this is the causal effect. What do you get?
**(b)** Now do what many dashboards do: run the same regression *only among visitors*.
What happens to the estimated ad effect, and why? (Draw the DAG --- which variable is
the collider?)

```{r problem1}
# your code here
```

# In-class problem 2: design your own experiment

Your product team wants to test a new checkout flow. The current completion rate is
60%, and a change would only be worth engineering effort if it moves completion by
at least 2 percentage points.

**(a)** How many users per arm do you need at $\alpha = 0.05$, power 0.8?
**(b)** The team says "we only have 5,000 users a week, and we want an answer in two
weeks." What minimum detectable effect does that traffic actually buy you (solve
`power.prop.test` for the detectable `p2` given n = 5000)?
**(c)** Two sentences for the product manager: what do you recommend, and what is the
risk of just running the underpowered test anyway?

```{r problem2}
# your code here
```
