---
title: "In-Class Session 7: LATE, Judges, Shift-Share, and Weak Instruments"
author: "Econometrics I"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4)
library(fixest)
set.seed(7)
```

This notebook reproduces the simulations behind Lecture 7 (the slides use `iv2_sims.R`) and
leaves you three problems. Only `fixest` is needed.

# Part 1: LATE in an encouragement design

A retailer emails a random half of its customers to nudge them onto its app. Twenty percent
would install it anyway (always-takers), forty percent install it only if emailed (compliers),
forty percent never do. The app's effect on spending differs by type.

```{r}
sim_late <- function(N = 10000, tau = c(always = 2.0, complier = 0.8, never = 0.5)) {
  type <- sample(names(tau), N, replace = TRUE, prob = c(0.2, 0.4, 0.4))
  Z <- rbinom(N, 1, 0.5)
  D <- as.integer(type == "always" | (type == "complier" & Z == 1))
  base <- c(always = 3, complier = 1, never = 0.5)[type]
  tenure <- c(always = 4, complier = 2, never = 1)[type] + rnorm(N)
  data.frame(y = base + tau[type] * D + rnorm(N), D, Z, tenure, type, tau = tau[type])
}
d <- sim_late()
etable(feols(y ~ D, d), feols(y ~ Z, d), feols(D ~ Z, d), feols(y ~ 1 | D ~ Z, d),
       headers = c("OLS", "reduced form", "first stage", "2SLS"))
c(LATE = mean(d$tau[d$type == "complier"]), ATE = mean(d$tau), ATT = mean(d$tau[d$D == 1]))
```

Who are the compliers? Abadie's trick: apply the Wald ratio to $X \cdot D$ instead of $y$.

```{r}
with(d, (mean(tenure[Z == 1] * D[Z == 1]) - mean(tenure[Z == 0] * D[Z == 0])) / (mean(D[Z == 1]) - mean(D[Z == 0])))
tapply(d$tenure, d$type, mean)    # the truth, which you never see in real data
```

# In-class problem 1: which rows move?

Re-simulate with the complier effect doubled (`tau = c(always = 2, complier = 1.6, never = 0.5)`),
then with the always-taker effect doubled instead. For each, say which of the four regression
rows change and which of the three truths (LATE, ATE, ATT) change, and why 2SLS never notices the
always-takers.

```{r problem1}
# your code here
```

# Part 2: A judge design with refund agents

Requests are routed at random to 50 agents who differ in generosity. The claim's merit $q$ is
seen by the agent, not by us; refunds build the most goodwill for weak claims.

```{r}
sim_refund <- function(N = 20000, J = 50, typed = FALSE) {
  q <- rnorm(N); agent <- sample(1:J, N, replace = TRUE)
  ctype <- sample(c("shipping", "damage"), N, replace = TRUE)
  gen <- rnorm(J, -0.5, 0.5)
  if (typed) { gen2 <- -0.5 - (gen + 0.5)                # agents lenient on shipping are strict on damage
    cutoff <- ifelse(ctype == "shipping", gen[agent], gen2[agent]) } else cutoff <- gen[agent]
  R <- as.integer(q + cutoff + rnorm(N, 0, 0.5) > 0)
  tau <- 20 - 15 * q
  d <- data.frame(spend = 100 + tau * R + 30 * q + rnorm(N, 0, 20), R, agent, q, tau, ctype)
  tot <- tapply(R, agent, sum); n_j <- tapply(R, agent, length)
  d$len <- (tot[agent] - R) / (n_j[agent] - 1)           # leave-one-out leniency
  d
}
dr <- sim_refund()
etable(feols(spend ~ R, dr, cluster = ~agent), feols(R ~ len, dr, cluster = ~agent),
       feols(spend ~ 1 | R ~ len, dr, cluster = ~agent), headers = c("OLS", "first stage", "2SLS"))
c(ATE = mean(dr$tau), effect_on_refunded = mean(dr$tau[dr$R == 1]))
```

The first-stage picture, and the monotonicity check by claim type:

```{r}
bins <- cut(dr$len, quantile(dr$len, seq(0, 1, .05)), include.lowest = TRUE)
plot(tapply(dr$len, bins, mean), tapply(dr$R, bins, mean), pch = 16, xlab = "agent leniency (leave-one-out)", ylab = "P(refund)")
sapply(split(dr, dr$ctype), function(s) coef(feols(R ~ len, s))[2])
```

# In-class problem 2: agents with tastes

Re-simulate with `typed = TRUE`: agents who are lenient on shipping claims are strict on damage
claims, and vice versa.

1. Re-run the overall first stage and the first stage by claim type. What do you see?
2. Re-run 2SLS with the pooled leniency measure. Is it still a LATE for anyone?
3. Fix it: compute leniency *within* claim type (each agent's leave-one-out rate on shipping
   claims, separately from damage claims) and re-estimate. Explain in two sentences what
   assumption the fix restores.

```{r problem2}
# your code here
```

# Part 3: Shift-share with Rotemberg weights

Markets differ in which sector their banks lend to; national sector credit shocks move local
credit growth; the question is the effect of local credit on firm entry (truth 0.4).

```{r}
sim_ss <- function(M = 1000, K = 6) {
  dom <- sample(1:K, M, replace = TRUE); sh <- matrix(0.4 / (K - 1), M, K); sh[cbind(1:M, dom)] <- 0.6
  g <- rnorm(K); eta <- rnorm(M); B <- as.vector(sh %*% g)
  x <- B + 0.5 * eta + rnorm(M, 0, 0.3); y <- 0.4 * x + eta + rnorm(M, 0, 0.3)
  list(d = data.frame(y, x, B), sh = sh, g = g)
}
s <- sim_ss(); d <- s$d
etable(feols(y ~ x, d), feols(y ~ 1 | x ~ B, d, vcov = "hetero"), headers = c("OLS", "2SLS"))
# Rotemberg weights: alpha_k proportional to g_k * sum_m s_mk * (x_m - mean x)
xt <- d$x - mean(d$x)
alpha <- sapply(1:6, function(k) s$g[k] * sum(s$sh[, k] * xt)); alpha <- alpha / sum(alpha)
beta_k <- sapply(1:6, function(k) coef(feols(y ~ 1 | x ~ z, data.frame(d, z = s$sh[, k] * s$g[k])))[2])
round(cbind(shock = s$g, weight = alpha, beta_k = beta_k), 2); sum(alpha * beta_k)
```

# In-class problem 3: drop the sector that carries the estimate

Remove the sector with the largest Rotemberg weight (set its shock to zero, or drop it from the
instrument) and re-estimate.

1. What happens to the 2SLS estimate and to the first-stage $F$?
2. If that sector's share-exogeneity claim were the doubtful one, what would you conclude from
   the comparison?
3. Bonus: Anderson-Rubin. For a grid of $\beta_0$ from 0 to 1, regress $y - \beta_0 x$ on $B$ and
   record whether the coefficient is significant at 5%. Report the set of $\beta_0$ not rejected and
   compare it to the 2SLS confidence interval.

```{r problem3}
# your code here
```
