---
title: "Hausman–Wu and Sargan (Hansen J) tests with fixest — R Notebook"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(dplyr)
library(fixest)
library(AER)       # for ivreg
library(sandwich)  # for robust variance calculations
library(Matrix)    # for matrix operations
set.seed(2025)
```

# Overview

This notebook simulates two scenarios and implements:
  
  * The matrix/formal Hausman test comparing OLS and IV estimates;
* The regression-based Wu/Hausman test (include first-stage residuals in the outcome);
* The Sargan test (homoskedastic version) and the heteroskedasticity-robust Hansen J statistic implemented from GMM moments.

We show both a **homogeneous** treatment-effect setting (where the tests behave as standard) and a **heterogeneous** treatment-effect setting (where Sargan/Hansen and Hausman can reject even with valid instruments).

---
  
  ## 1. Simulate data — two instruments, possible heterogeneity
  
  We simulate a setting where there are two instruments `z1` and `z2`. Each instrument affects `x` only for a different subgroup (defined by `w`), and treatment effects differ across those subgroups in the heterogeneous scenario.

```{r sim-data}
n <- 2000
# covariate that determines subgroup
w <- rbinom(n, 1, 0.5)

# two instruments (binary), independent
z1 <- rbinom(n, 1, 0.5)
z2 <- rbinom(n, 1, 0.5)

# reduced-form noise
v <- rnorm(n)
u <- rnorm(n)

# instrument relevance differs by subgroup: z1 matters when w==0, z2 when w==1
gamma1 <- 1.0
gamma2 <- 1.0
x <- gamma1 * z1 * (1 - w) + gamma2 * z2 * w + 0.5 * v

# Errors correlated with x (endogeneity): let u correlate with v
u <- 0.5 * v + rnorm(n, sd = 1)

# Homogeneous effect
beta_hom <- 2
y_hom <- beta_hom * x + rnorm(n)

# Heterogeneous effect: different beta by subgroup
beta0 <- 1.0
beta1 <- 3.0
# Here beta differs by w
y_het <- x * (beta0 * (1 - w) + beta1 * w) + rnorm(n)

# Put into data.frame
df <- data.frame(y_hom, y_het, x, w, z1, z2)
head(df)
```

---
  
  ## 2. OLS and IV estimates
  
  We estimate OLS with `fixest::feols` and IV with `AER::ivreg`. For IV we use both instruments `z1` and `z2` together.

```{r ols-iv}
# OLS (fixest)
ols_hom <- feols(y_hom ~ x, data = df)
ols_het <- feols(y_het ~ x, data = df)

# IV (AER::ivreg) — instrument x by z1 + z2
iv_hom <- feols(y_hom ~ 1 | x~ z1 + z2, data = df)
iv_het <- feols(y_het ~ 1 | x~z1 + z2, data = df)

summary(ols_hom)
summary(iv_hom)
summary(ols_het)
summary(iv_het)
etable(ols_hom,ols_het,iv_hom,iv_het)
```

Take note: in the homogeneous case OLS and IV should be closer (depending on endogeneity), while in the heterogeneous case the IV identifies a complier-weighted average that can differ from OLS.

---
  
  ## 3. Hausman–Wu (matrix) test — difference in estimators
  
  We implement the matrix-form Hausman statistic

( H = (\hat{\beta}*{IV} - \hat{\beta}*{OLS})' [Var(\hat{\beta}*{IV}) - Var(\hat{\beta}*{OLS})]^{-1} (\hat{\beta}*{IV} - \hat{\beta}*{OLS}) )

(Here models are just- or over-identified but the test is valid for K parameters.)

```{r hausman-matrix}
# Extract estimates and covariance matrices
beta_ols_hom <- coef(ols_hom)["x"]
beta_iv_hom  <- coef(iv_hom)["fit_x"]

# Variances: use sandwich robust cov for IV and OLS
vcov_ols_hom <- vcov(ols_hom, vcov = sandwich)
vcov_iv_hom  <- sandwich::vcovHC(iv_hom, type = "HC0")

# For single-parameter x case, forms simplify to scalar
delta_hom <- beta_iv_hom - beta_ols_hom
var_diff_hom <- as.numeric(vcov_iv_hom["fit_x","fit_x"] - vcov_ols_hom["x","x"])
H_hom <- (delta_hom^2) / var_diff_hom
pval_hom <- 1 - pchisq(H_hom, df = 1)

# Repeat for heterogeneous outcome
beta_ols_het <- coef(ols_het)["x"]
beta_iv_het  <- coef(iv_het)["fit_x"]
vcov_ols_het <- vcov(ols_het, vcov = sandwich)
vcov_iv_het  <- sandwich::vcovHC(iv_het, type = "HC0")
delta_het <- beta_iv_het - beta_ols_het
var_diff_het <- as.numeric(vcov_iv_het["fit_x","fit_x"] - vcov_ols_het["x","x"])
H_het <- (delta_het^2) / var_diff_het
pval_het <- 1 - pchisq(H_het, df = 1)

data.frame(
  scenario = c("homogeneous","heterogeneous"),
  beta_ols = c(beta_ols_hom, beta_ols_het),
  beta_iv  = c(beta_iv_hom, beta_iv_het),
  H_stat   = c(H_hom, H_het),
  p_value  = c(pval_hom, pval_het)
)
```

> Note: we used heteroskedasticity-robust variances for both estimators when forming the difference.

---

## 4. Wu/Hausman regression-based test (first-stage residuals)

The Wu test regresses `y` on `x` and the first-stage residuals `v_hat` from the regression of `x` on instruments. Under exogeneity the coefficient on `v_hat` should be zero.

```{r wu-test}
# First-stage: regress x on instruments (separately for hom/het design; same X used)
first_stage <- feols(x ~ z1 + z2, data = df)
df$vhat <- resid(first_stage)

# Augmented outcome regression
aug_hom <- feols(y_hom ~ x + vhat, data = df)
aug_het <- feols(y_het ~ x + vhat, data = df)

summary(aug_hom)
summary(aug_het)

# Test whether coefficient on vhat is zero (t-test / F-test)
coeftable(aug_hom)
coeftable(aug_het)
```

If the coefficient on `vhat` is significant, that's evidence of endogeneity (reject exogeneity).

---
  
  ## 5. Sargan test (homoskedastic) and Hansen J (robust)
  
  We compute the homoskedastic Sargan statistic as:
  
  ( J = n \cdot R^2 )

from regressing IV residuals on all instruments. For the heteroskedasticity-robust Hansen J, we implement the GMM-style statistic

( J = n , \hat{g}(\hat{\beta})' \hat{\Omega}^{-1} \hat{g}(\hat{\beta}) )

where ( \hat{g} = \frac{1}{n} Z'\hat{u} ) and ( \hat{\Omega} = \frac{1}{n} Z' \hat{u} \hat{u}' Z ).

```{r sargan-hansen}
# IV residuals
u_iv_hom <- resid(iv_hom)
u_iv_het <- resid(iv_het)

# Instrument matrix
Z <- as.matrix(cbind(1, df$z1, df$z2)) # include intercept (we'll account for dims)
n <- nrow(df)
K <- 1  # number of endogenous regressors (x only)
L <- ncol(Z) - 1 # instruments count excluding intercept

# -- Homoskedastic Sargan (classical) --
# regress residuals on instruments (exclude intercept for regressing)
sargan_lm_hom <- lm(u_iv_hom ~ z1 + z2, data = df)
J_hom_classic <- n * summary(sargan_lm_hom)$r.squared
p_sargan_hom <- 1 - pchisq(J_hom_classic, df = L - K)

sargan_lm_het <- lm(u_iv_het ~ z1 + z2, data = df)
J_het_classic <- n * summary(sargan_lm_het)$r.squared
p_sargan_het <- 1 - pchisq(J_het_classic, df = L - K)

# -- Hansen J (robust) via GMM moment approach --
# helper to compute robust J
hansen_J <- function(u, Z) {
  n <- length(u)
  Zmat <- as.matrix(Z)
  g_hat <- drop(t(Zmat) %*% u) / n          # vector of length dim(Z)
  # Omega_hat = (1/n) Z' (u u') Z = (1/n) * t(Z * u) %*% (Z * u)
  Zu <- Zmat * as.numeric(u)
  Omega_hat <- crossprod(Zu) / n
  # Remove intercept row/col if present — use full Z including intercept here but
  # we will compute J over instrument rows excluding intercept if desired.
  # We'll use all instruments (z1,z2) rows only (exclude intercept component)
  # Identify indices corresponding to z1,z2 in Zmat: here 2 and 3
  idx <- 2:ncol(Zmat)
  g_sub <- g_hat[idx]
  Omega_sub <- Omega_hat[idx, idx]
  J <- n * as.numeric(t(g_sub) %*% solve(Omega_sub) %*% g_sub)
  df_j <- length(idx) - K
  list(J = J, df = df_j, pval = 1 - pchisq(J, df_j))
}

# Build Z without intercept for moments
Z_noint <- cbind(df$z1, df$z2)

h_hom <- hansen_J(u_iv_hom, cbind(1, Z_noint))
h_het <- hansen_J(u_iv_het, cbind(1, Z_noint))

# Show results
res_sargan <- data.frame(
  scenario = c("homogeneous","heterogeneous"),
  J_classic = c(J_hom_classic, J_het_classic),
  p_classic = c(p_sargan_hom, p_sargan_het),
  J_hansen = c(h_hom$J, h_het$J),
  p_hansen = c(h_hom$pval, h_het$pval)
)
res_sargan
```

Interpretation:
  
  * If `p` is small (e.g. < 0.05) we reject the null that all instruments are valid.
* In the heterogeneous case, we may observe rejection even when instruments are 'valid' in the sense that each instrument affects `x` exogenously for its subgroup — because different instruments identify distinct local effects.

---
  
  ## 6. Wrap-up / notes
  
  * The notebook used `fixest` for OLS and first-stage regressions; IV was computed with `AER::ivreg` for convenience.
* We computed both the matrix Hausman statistic and the regression-based Wu test — both are valid ways to probe endogeneity under homogeneous-effect assumptions.
* We computed both the classical Sargan (homoskedastic) statistic and a heteroskedasticity-robust Hansen J using a GMM moment approach.

---
  
```{r session-info}
sessioninfo::session_info()
```
