---
title: "Problem Set "
author: "Prof. Conlon"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


\newcommand{\E}[1]{\ensuremath{\mathbb{E}\big[#1\big]}}
\newcommand{\Emeasure}[2]{\ensuremath{\mathbb{E}_{#1}\big[#2\big]}}

## Packages to Install


**The packages used this week are**

* ggplot2

```{r,comment='\t\t',echo=FALSE}

## This is a code chunk: it is outlined in grey and has R code inside of it
## Note: you can change what is shown in the final .pdf document using arguments 
##       inside the curly braces at the top {r, comment='\t\t'}. For example, you 
##       can turn off print statements showing in the .pdf by adding 'echo=False' 
##       i.e. changing the header to {r, comment='\t\t',echo=FALSE}

## ~~~~~~~~~~~~~~ CODE SETUP ~~~~~~~~~~~~~~ ##

# ~ This bit of code will be hidden after Problem Set 1 ~
#
# This section sets up the correct directory structure so that
#  the working directory for your code is always in the data folder

# Retrieve the code working directory
#script_dir = dirname(sys.frame(1)$ofile)
initial_options <- commandArgs(trailingOnly = FALSE)
render_command <- initial_options[grep('render',initial_options)]
script_name <- gsub("'", "", 
                    regmatches(render_command, 
                               gregexpr("'([^']*)'",
                               render_command))[[1]][1])

# Determine OS (backslash versus forward slash directory system)
sep_vals = c(length(grep('\\\\',script_name))>0,length(grep('/',script_name))>0)
file_sep = c("\\\\","/")[sep_vals]
```


## Problem 1 (Coding Exercise)

This exercise asks you to implement and assess the performance of the bootstrap for the linear regression model. Suppose you have the linear regression model:

\begin{align*}
  y_{i} = \beta_{0} + \beta_{1}x_{i} + \epsilon_{i}
\end{align*}

where,

\begin{itemize}
  \item[-] $x_{i} \sim U[0,2]$
  \item[-] $\epsilon_{i} \vert x_{i} \sim U[-1,1]$
  \item[-] $\beta_{0} = \beta_{1} = 1$ 
\end{itemize}

We ask you to answer the following questions:

\begin{itemize}
\item[a.] Write a code that generates i.i.d. samples of sizes $n=10,50,200$ from that distribution, computes (1) the least squares estimator for $\beta$, (2) the t-ratio for the least squares coefficient $\beta_{1}$, $t_{n} = \frac{\hat{\beta}_{1,LS} - 1}{\hat{s.e.}(\hat{\beta}_{1,LS})}$, and (3) the least square residuals $\hat{\epsilon}_{i} = y_{i} - \hat{\beta}_{0,LS} - \hat{\beta}_{1,LS} x_{i}$

\item[b.] Write a code for drawing $n$ times at random from the discrete uniform distribution over the estimated residuals $\hat{\epsilon}_{1},...,\hat{\epsilon}_{n}$ (i.e. with replacement).

\item[c.] Use your code from parts (a) and (b) to implement the residual bootstrap - assuming that $\epsilon_{i}$ and $x_{i}$ are independent - to estimate the 95th percentiles of the respective distributions of $\hat{\beta}_{1,LS}$ and $t_{n}$

\item[d.] Repeat part (a) for sample size $n=10,50,200$ with $200$ replications, where you keep the initial draws of $x_{1},..,x_{n}$ from part (a) and only generate new residuals from their conditional distribution. Compute $\hat{\beta}_{1,LS}$ and the statistic $t_{n}$ using $200$ independent samples of size $n$. Use your results to compute a simulated estimate for the 95th percentiles of the respective sampling distributions for $\hat{\beta}_{1,LS}$ and $t_{n}$.

\item[e.] Compare your results from (c) and (d). What do you conclude about the performance of the bootstrap? How does it compare to the 95th percentile of the asymptotic distribution of $t_{n}$?

\item[f.] See if you can construct a subsampled confidence interval for $n=200$ with $a_n = 25$. How does it compare to the bootstrapped CI?

\end{itemize}

