---
title: 'Exercises: Week 2'
author: "Prof. Conlon"
date: 'Due: 2/8/21'
output:
  pdf_document: default
  html_notebook: default
---

1. Let's load the Boston HMDA data.

The function should take the following arguments:

  * dir: debt to income ratio
  * hir : housing to income ratio
  * single : dummy for single borrower
  * self : dummy for self-employed

 
```{r}
library("Ecdat")
data("Hmda")

probit <- glm(deny ~ dir + hir + single + self, data = Hmda, family = binomial(link = "probit"))
logit <- glm(deny ~ dir + hir + single + self, data = Hmda, family = binomial(link = "logit"))

summary(logit)
```

2. Consider the regression model of the logit regression:
$$
deny_i = F(\beta_1\cdot dir_i + \beta_2\cdot hir_i + \beta_3\cdot single_i + \beta_4 \cdot self_i)
$$
For a single observation compute the contribution to the log-likelihood (analytically)

3. For a single observation compute the Score (analytically).

4. Compute the Hessian Matrix and Fisher information (analytically).

5. Code up the Fisher Information for the logit model above $I(\widehat{\beta})$ using the Hessian Matrix.

6. Code up the Fisher Information for the logit model above $I(\widehat{\beta})$ using the score method.

7. Compute the standard errors from the Fisher information and compare them to the standard errors reported from the regression. How do they compare?

8. Generate $n=100$ observations where $\lambda = 15$ from a poisson model:

$$ Y_i \sim Pois(\lambda)$$

9. The poisson distribution is a discrete distribution for count data where the p.m.f. is given by:

$$Pr(Y_i =k )=\frac{\lambda^k e^{-\lambda}}{k!}a$$

10. Write the log-likelihood $\ell(y_1,\ldots,y_n; \lambda)$ (analytically).

11. Write the Score contribution $\mathcal{S}_i(y_i; \lambda)$ (analytically).

12. Write the Hessian Contribution $\mathcal{H}_i(y_i; \lambda)$ (analytically).

13. Code up the log-likelihood function

```{r}
pois_log_lik <- function(lambda,y){
  
  return(ll)
}
```

14. Find the value of $\lambda$ that maximizes your log likelihood using \texttt{optim} in R.

15. Write a function that returns the standard error of $\hat\lambda$:

```{r}
pois_se <- function(lambda_hat,y){
  
  return(se)
}