---
title: "In-Class Session 8: Panel Data, Fixed Effects, and Event Studies"
author: "Econometrics I"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4)
library(fixest)
library(data.table)
set.seed(8)
```

This notebook reproduces the simulations behind Lectures 8a and 8b (the slides use
`panel_sims.R`, which has the same code with more plotting) and leaves you four short
problems. Everything is simulated, so you always know the right answer.

# Part 1: Panel data and fixed effects

## Reproduce: the management-practices panel

Firm quality $\alpha_i$ raises productivity *and* makes firms adopt better practices.
The true effect of practices on log productivity is $\beta = 0.10$.

```{r}
beta_true <- 0.10
sim_panel <- function(N = 200, T = 8, rho = 0.7, sd_alpha = 0.5, sd_eps = 0.3, me_sd = 0) {
  alpha <- rnorm(N, 0, sd_alpha)
  gamma <- rnorm(T, 0, 0.1)
  d <- CJ(firm = 1:N, year = 1:T)
  d[, alpha := alpha[firm]]
  d[, x := rho * alpha + sqrt(1 - rho^2) * (rnorm(N)[firm] * 0.6) + rnorm(.N, 0, 0.8)]
  d[, y := 1 + beta_true * x + alpha + gamma[year] + rnorm(.N, 0, sd_eps)]
  if (me_sd > 0) d[, x := x + rnorm(.N, 0, me_sd)]
  d[]
}
d <- sim_panel()
```

Pooled OLS, the within (FE) estimator, and first differences:

```{r}
fit_pooled <- feols(y ~ x, d, cluster = ~firm)
fit_fe     <- feols(y ~ x | firm + year, d, cluster = ~firm)
fit_fd     <- feols(d(y) ~ d(x), d, panel.id = ~firm + year, cluster = ~firm)
etable(fit_pooled, fit_fe, fit_fd, headers = c("pooled", "within", "first diff"))
```

## Reproduce: random effects by hand

Random effects is GLS with equicorrelated errors, which is OLS after *quasi*-demeaning
with weight $\theta = 1 - \sqrt{\sigma^2_\varepsilon / (\sigma^2_\varepsilon + T\sigma^2_\alpha)}$.
Estimate the two variances from the within and between regressions (Swamy--Arora), then transform.

```{r}
re_fit <- function(d) {
  within <- feols(y ~ x | firm, d)
  s2_eps <- sum(resid(within)^2) / (nrow(d) - uniqueN(d$firm) - 1)
  b <- d[, .(ybar = mean(y), xbar = mean(x), T_i = .N), by = firm]
  between <- lm(ybar ~ xbar, b)
  s2_alpha <- max(0, sum(resid(between)^2) / between$df.residual - s2_eps / mean(b$T_i))
  d <- merge(d, b, by = "firm")
  d[, theta := 1 - sqrt(s2_eps / (s2_eps + T_i * s2_alpha))]
  fit <- lm(I(y - theta * ybar) ~ I(x - theta * xbar), d)
  list(beta = unname(coef(fit)[2]), se = unname(sqrt(diag(vcov(fit)))[2]),
       theta = mean(d$theta), s2_eps = s2_eps, s2_alpha = s2_alpha)
}
re <- re_fit(d)
str(re)
```

Compare $\hat\theta$ to the formula: with $T = 8$ and $\sigma^2_\alpha / \sigma^2_\varepsilon \approx$
`r round(re$s2_alpha / re$s2_eps, 2)`, RE is already most of the way to FE. Yet it is biased,
because $\alpha_i$ is correlated with $x$.

## Reproduce: Mundlak (correlated random effects)

Add the firm mean of $x$ as a regressor. The coefficient on $x_{it}$ is the FE estimate; the
coefficient on $\bar x_i$ is a cluster-robust Hausman test.

```{r}
d[, xbar := mean(x), by = firm]
fit_mund <- feols(y ~ x + xbar | year, d, cluster = ~firm)
etable(fit_fe, fit_mund, headers = c("within", "Mundlak"))
```

## Reproduce: partial pooling of store effects

Sixty stores with very different amounts of data. The partial-pooling estimate shrinks each
store's mean toward the grand mean by a weight equal to its signal-to-noise ratio.

```{r}
S <- 60; alpha <- rnorm(S, 0, 0.4); n_s <- sample(c(4, 8, 16, 32, 64), S, replace = TRUE)
ybar <- alpha + rnorm(S, 0, 1 / sqrt(n_s)); grand <- mean(ybar)
s2_alpha <- max(0, var(ybar) - mean(1 / n_s))                     # method of moments, sigma_eps^2 = 1 known
w <- s2_alpha / (s2_alpha + 1 / n_s); eb <- grand + w * (ybar - grand)
c(no_pooling = mean((ybar - alpha)^2), complete = mean((grand - alpha)^2), partial = mean((eb - alpha)^2))
rbind(top5_by_mean_n = n_s[order(-ybar)][1:5], top5_by_partial_n = n_s[order(-eb)][1:5])
```

# In-class problem 1: turn the correlation off and on

Re-simulate with `rho = 0` (no correlation between quality and practices) and with
`rho = 0.9`. For each, report pooled, within, and RE estimates. Which estimator moves,
and which stays put? Then explain in one sentence what the RE assumption is buying you
when `rho = 0`.

```{r}
# your code here
```

# In-class problem 2: measurement error

Re-simulate with `rho = 0` and `me_sd = 0.5` (noise added to the observed practices score).
Now pooled OLS has no omitted-variable problem. Compare pooled and within estimates to
$0.10$. Which is more attenuated, and why does demeaning make measurement error worse?
(Hint: what does demeaning do to the *signal* variance of $x$ versus the *noise* variance?)

```{r}
# your code here
```

# Part 2: Event studies

## Reproduce: the single-timing event study

400 firms, 24 quarters. Half adopt a pricing tool in quarter 13; the effect ramps from
0.0125 to 0.05 over four quarters. Never-treated firms get the reference value of event time
so they stay in the regression and identify the period effects.

```{r}
tau_true <- function(k) ifelse(k >= 0, 0.05 * pmin(1, (k + 1) / 4), 0)
sim_es <- function(N = 400, T = 24, E0 = 13, trend = 0, anticipate = 0, sd_eps = 0.08) {
  d <- CJ(firm = 1:N, t = 1:T)
  d[, treated := as.integer(firm <= N / 2)]
  d[, alpha := rnorm(N, 0, 0.3)[firm]]
  gam <- cumsum(rnorm(T, 0.005, 0.02))
  d[, k := ifelse(treated == 1, t - E0, -1L)]          # controls carry the reference value
  d[, tau := ifelse(treated == 1, tau_true(t - E0), 0)]
  if (anticipate > 0) d[treated == 1 & (t - E0) < 0 & (t - E0) >= -anticipate, tau := 0.0125]
  d[, y := alpha + gam[t] + tau + trend * treated * (t - E0) + rnorm(.N, 0, sd_eps)]
  d[]
}
es <- sim_es()
fit_es <- feols(y ~ i(k, ref = -1) | firm + t, es, cluster = ~firm)
iplot(fit_es, ref.line = -0.5, main = "Event study: log gross margin around adoption")
lines(-12:11, tau_true(-12:11), col = "red", lwd = 2)
```

The joint test on the pre-period coefficients, and the static two-way FE coefficient for comparison:

```{r}
wald(fit_es, keep = "k::-")
es[, post := treated * (t >= 13)]
coef(feols(y ~ post | firm + t, es, cluster = ~firm))
mean(tau_true(0:11))    # what the static coefficient is averaging
```

# In-class problem 3: anticipation

Re-simulate with `anticipate = 2`: firms start adjusting two quarters before the tool goes live.

1. Re-estimate and plot. What happens to the pre-period coefficients, and what happens to
   the *post*-period coefficients even though the true post effects have not changed?
2. Fix it by changing the reference period to `k = -3` (`ref = -3`). Confirm the post-period
   estimates are back on the red line.
3. Bonus: instead of moving the reference, drop quarters $k = -2, -1$ from the sample. Does it
   work? What did you give up?

```{r}
# your code here
```

# Part 3: The returns event study

## Reproduce: earnings announcements and CARs

200 firms, one announcement each, 261 trading days per firm. The market model is estimated on
days $-250$ to $-11$; abnormal returns are computed on $-10$ to $+10$. A one-SD earnings
surprise moves the stock 2% on day 0 and drifts another 1% over the next ten days.

```{r}
N_c <- 200; days <- -250:10
S <- rnorm(N_c)
car <- CJ(firm = 1:N_c, day = days)
car[, `:=`(S = S[firm], beta = rnorm(N_c, 1, 0.3)[firm], a = rnorm(N_c, 0, 2e-4)[firm],
           Rm = rnorm(.N, 4e-4, 0.010))]        # each firm on its own calendar dates
car[, ar_true := fifelse(day == 0, 0.02 * S, fifelse(day >= 1 & day <= 10, 0.001 * S, 0))]
car[, R := a + beta * Rm + ar_true + rnorm(.N, 0, 0.02)]

est <- car[day <= -11, { f <- lm(R ~ Rm); .(a_hat = coef(f)[1], b_hat = coef(f)[2], s_hat = summary(f)$sigma) }, by = firm]
car <- merge(car, est, by = "firm")
car[, AR := R - a_hat - b_hat * Rm]
ev <- car[day >= -10 & day <= 10][, CAR := cumsum(AR), by = firm]
```

Average abnormal returns by day, split by the sign of the surprise:

```{r}
agg <- ev[, .(AAR = mean(AR), CAAR = mean(CAR)), by = .(pos = S > 0, day)][order(pos, day)]
plot(agg[pos == TRUE, day], 100 * agg[pos == TRUE, CAAR], type = "b", pch = 16, col = "navy",
     ylim = 100 * range(agg$CAAR), xlab = "event day", ylab = "CAAR (%)")
lines(agg[pos == FALSE, day], 100 * agg[pos == FALSE, CAAR], type = "b", pch = 16, col = "firebrick")
abline(h = 0, lty = 3); abline(v = 0, lty = 2)
legend("topleft", bty = "n", legend = c("positive surprise", "negative surprise"), col = c("navy", "firebrick"), lwd = 1)
```

CARs over windows, with the cross-sectional $t$-statistic, and the earnings response coefficient:

```{r}
win <- function(k1, k2) ev[day >= k1 & day <= k2, .(CAR = sum(AR), s_hat = s_hat[1], S = S[1]), by = firm]
for (w in list(c(-10, -2), c(-1, 1), c(0, 10))) {
  z <- win(w[1], w[2])
  cat(sprintf("window [%3d,%3d]: CAAR = %6.2f%%  t = %5.2f   (S>0: %5.2f%%, S<0: %5.2f%%)\n",
              w[1], w[2], 100 * mean(z$CAR), mean(z$CAR) / (sd(z$CAR) / sqrt(nrow(z))),
              100 * mean(z[S > 0, CAR]), 100 * mean(z[S < 0, CAR])))
}
feols(CAR ~ S, win(0, 10), vcov = "hetero")     # truth: slope 0.03
```

# In-class problem 4: clustered event dates

Every firm above announced on a different calendar day, so their abnormal returns are
independent. Now suppose all 200 firms announce on the *same* day (think: earnings season
compressed into one date, or a regulation that hits every bank at once). Simulate this by
drawing **one** market-return series and giving it to every firm (replace the per-row `Rm`
draw with a per-day draw shared across firms), and add a common shock of $-1\%$ to every
firm's return on day $0$ that has nothing to do with earnings.

1. Recompute the window $[0, 0]$ CAAR and its cross-sectional $t$-statistic for the full sample.
   What does the test conclude, and is it right?
2. Now regress $\text{CAR}_i(0, 0)$ on $S_i$. Is the *slope* still estimated correctly?
   Explain why the ERC survives what the average CAR does not.
3. What would you do differently in a real study where events cluster in time?

```{r}
# your code here
```
