---
title: "Problem Set 6"
author: "Prof. Conlon"
date: "Due: 5/5/20"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


\newcommand{\E}[1]{\ensuremath{\mathbb{E}\big[#1\big]}}
\newcommand{\Emeasure}[2]{\ensuremath{\mathbb{E}_{#1}\big[#2\big]}}

## Packages to Install


**The packages used this week are**

* ggplot2
* xtable (build tables quickly)
* data.table (data tables are computationally efficient and IMHO easier to work with)


```{r,comment='\t\t',echo=FALSE}

## This is a code chunk: it is outlined in grey and has R code inside of it
## Note: you can change what is shown in the final .pdf document using arguments 
##       inside the curly braces at the top {r, comment='\t\t'}. For example, you 
##       can turn off print statements showing in the .pdf by adding 'echo=False' 
##       i.e. changing the header to {r, comment='\t\t',echo=FALSE}

## ~~~~~~~~~~~~~~ CODE SETUP ~~~~~~~~~~~~~~ ##

# ~ This bit of code will be hidden after Problem Set 1 ~
#
# This section sets up the correct directory structure so that
#  the working directory for your code is always in the data folder

# Retrieve the code working directory
#script_dir = dirname(sys.frame(1)$ofile)
initial_options <- commandArgs(trailingOnly = FALSE)
render_command <- initial_options[grep('render',initial_options)]
script_name <- gsub("'", "", 
                    regmatches(render_command, 
                               gregexpr("'([^']*)'",
                               render_command))[[1]][1])

# Determine OS (backslash versus forward slash directory system)
sep_vals = c(length(grep('\\\\',script_name))>0,length(grep('/',script_name))>0)
file_sep = c("\\\\","/")[sep_vals]

# Get data directory
split_str = strsplit(script_name,file_sep)[[1]]
len_split = length(split_str) - 2
data_dir = paste(c(split_str[1:len_split],'data',''),collapse=file_sep)

# Check that the data directory contains the files for this weeks assignment
data_files = list.files(data_dir)
if(!('lending_club_07_to_11.csv' %in% data_files)){
  cat("ERROR: DATA DIRECTORY NOT CORRECT\n")
  cat(paste("data_dir variable set to: ",data_dir,collapse=""))
}

```


## Problem 1 (Coding Exercise: Next Year)

Discrete Choice Question Here

## Problem 2 (Coding Exercise)

This exercise will walk you through a prediction task. I have downloaded data from a peer-to-peer lending platform, Lending Club. The dataset you will work with is: __lending_club_07_to_11_cleaned.csv__. Lending Club provides detailed characteristic information regarding loans, both information on the borrower, as well as, the loan itself. Your goal will be to build a model to predict the outcome of a loan, i.e. whether an individual paid off a loan or did not pay off a loan. In our case, a good outcome is if the loan is fully paid off, a bad outcome is if the loan is charged off. 

The target variable for the analysis is **loan_status**, where:

\begin{align*}
  loan\_status = 
    \begin{cases}
      1 \mbox{ if loan is paid off } \\
      0 \mbox{ if loan is not paid off }
    \end{cases}
\end{align*}

\begin{itemize}

\item[a.] This is going to be a more DIY style exercise, provide a list of the variables you plan to use for the analysis. Give a short discussion for why you excluded other variables.

\item[b.] Regularization is an important step when using an machine learning algorithm, regularzie the variables that you have included. Briefly, why is regularization important?

\item[c.] Provide a simple correlational table to give you a sense of the relationship between your covariates. Do you notice any interesting patterns?

\item[d.] Split the dataset into a single test and training set, a simple rule of thumb is an 40/60 split. How did you build these two sets?

\item[e.] Using your training set, run a logistic regression, a random forest, and a gradient boosted random forest. To show your results, present both a measure of miscalssification error, accuracy and a confusion matrix.

\item[f.] One easy way to improve model performance is cross-validation. Do a k-fold cross validation, where k=5, using the best performing model from part (e.). Re-report the misclassification error, accuracy and a confusion matrix. 

\end{itemize}